From Reversible Programs to \\ Univalent Universes and Back

### 2021-07-09 Pi+/paper/reversible1.tex
intro and level 2

To see how this language expresses reversible circuits, we present a few examples. First it is possible to directly
mimic the \verb|qasm|-perspective by defining types that describe sequences of booleans. We use the type
$\mathbb{2} = \onet + \onet$ to represent booleans with $\inlv{\Acon{tt}}$ representing \textsf{true} and
$\inrv{\Acon{tt}}$ representing $\textsf{false}$. Boolean negation (the \verb|x|-gate) is straightforward to define using
the primitive combinator $\swapp$. We can represent $n$-bit words using an n-ary product of boolean values, thus the
type $\mathbb{2} \times (\mathbb{2} \times \mathbb{2})$ (abbreviated $\mathbb{B}~3$) corresponds to a collection of
wires that can transmit three bits.
%
% For example, we can express a 3-bit word reversal operation as follows:
% $\Afun{reverse} : \mathbb{B}~3 \iso \mathbb{B}~3$
% $\Afun{reverse} = \swapt \fatsemi (\swapt  \otimes  \idc)~ \fatsemi \assocrt$
% \noindent The manual trace of $\Afun{reverse}$ below confirms that it indeed reverses the three bits:
% \[\begin{array}{rlr}
%                          & (v_1, (v_2, v_3)) \\
%     \swapt               & ((v_2, v_3), v_1) \\
%     \swapt \otimes  \idc & ((v_3, v_2), v_1) \\
%     \assocrt             & (v_3, (v_2, v_1)) \\
%   \end{array}\]
%subcode source isomorphisms.tex:979
%
To express the \verb|cx| and \verb|ccx| gates, we need to encode a notion of conditional expression. Such conditionals
turn out to be expressible using the distributivity and factoring identities of rigs as shown below:

@@

All of this is only half the story, however. A sound semantics for $\PiLang$ in weak rig
groupoids was established by \citet{caretteComputingSemiringsWeak2016}, and conjectured to be complete. For this
semantics, rig groupoid \emph{coherence conditions} that identify different syntactic representations of the same
permutation~\cite{laplaza72,Carette2016} were collected  in a
second level of $\PiLang$ syntax as level-2 combinators.  Each level-2 combinator is of the form $c_1 \isotwo c_2$ for appropriate
$c_1$ and $c_2$ of the same level-1 type $A \isoone B$ and asserts that $c_1$ and $c_2$ denote the same bijection. For example, we have:

@@

\combtwo{}

\noindent where the first level-2 combinator states the obvious fact that compositions with the identity can be
optimized and the second level-2 combinator states the more interesting observation that swapping the two sides of a sum
type and then applying $c_1$ to the left and $c_2$ to the right is equivalent to applying $c_2$ to the left and $c_1$ to
the right and then swapping. The full set of level-2 combinators is large and is only listed in the accompanying code.

>>>>

To see how this language expresses reversible circuits, we present a few examples in the Agda embedding of the
language. The Agda syntax is almost identical to the syntax in Fig.~\ref{fig:pi-terms} except that sequential
composition is written using $\circledcirc$. First it is possible to directly mimic the \verb|qasm|-perspective by
defining types that describe sequences of booleans. We use the type $\mathbb{2} = \onet + \onet$ to represent booleans
with $\inlv{\Acon{tt}}$ representing \textsf{true} and $\inrv{\Acon{tt}}$ representing $\textsf{false}$. Boolean
negation (the \verb|x|-gate) is straightforward to define using the primitive combinator $\swapp$. We can represent
$n$-bit words using an n-ary product of boolean values, thus the type $\mathbb{2} \times (\mathbb{2} \times \mathbb{2})$
(abbreviated $\mathbb{B}~3$) corresponds to a collection of wires that can transmit three bits.
%
% For example, we can express a 3-bit word reversal operation as follows:
% $\Afun{reverse} : \mathbb{B}~3 \iso \mathbb{B}~3$
% $\Afun{reverse} = \swapt \fatsemi (\swapt  \otimes  \idc)~ \fatsemi \assocrt$
% \noindent The manual trace of $\Afun{reverse}$ below confirms that it indeed reverses the three bits:
% \[\begin{array}{rlr}
%                          & (v_1, (v_2, v_3)) \\
%     \swapt               & ((v_2, v_3), v_1) \\
%     \swapt \otimes  \idc & ((v_3, v_2), v_1) \\
%     \assocrt             & (v_3, (v_2, v_1)) \\
%   \end{array}\]
%subcode source isomorphisms.tex:979
%
To express the \verb|cx| and \verb|ccx| gates, we need to encode a notion of conditional expression. Such conditionals
turn out to be expressible using the distributivity and factoring identities of rigs as shown below:

@@

All of this is only half the story, however. A sound semantics for $\PiLang$ in weak rig
groupoids was established by \citet{caretteComputingSemiringsWeak2016}, and conjectured to be complete. For this
semantics, rig groupoid \emph{coherence conditions} that identify different syntactic representations of the same
permutation~\cite{laplaza72,Carette2016} were collected  in a
second level of $\PiLang$ syntax as level-2 combinators.  Each level-2 combinator is of the form $c_1 \isotwo c_2$ for appropriate
$c_1$ and $c_2$ of the same level-1 type $A \isoone B$ and asserts that $c_1$ and $c_2$ denote the same bijection. For
example, we have the following level-2 combinators dealing with associativity:

@@

\leveltwoblockone{}

\noindent The full set of level-2 combinators is large; the remaining combinators are listed in App.~\ref{app:leveltwo}.

\begin{toappendix}
  \label{app:leveltwo}

  The additional level-2 combinators:

  \medskip
  \leveltwoblocktwo{}
\end{toappendix}
